Debugging by \textit{lastChange}

### 2010-10-15 papers/pldi2011/sigplanconf-template.tex

As demonstrated in Figure ~\ref{fig:example-points}, the developer examinated three points of execution. These points-the history of the search for the defect-are available through the debugger's interface. On the top of the left panel in Figure ~\ref{fig:example3} there is an opened list which shows all three examinated points. The first one is the breakpoint on line 22, the second one is the point which is the last change of \texttt{myObject.myProperty} before the reaching the breakpoint and finally the last one is the point of execution in which \texttt{myCondition.value} gets the \texttt{undefined} value. Moreover, the source lines related to these points are tagged with red-cycle icons.

Notice that in our example, \textit{lastChange} combines some aspects of breakpoint and of log-based debugging. Like breakpoint debugging, the developer re-executes a live runtime without changing the source and without a special execution environment beyond the debugger. The state of the program memory and the call stack are available at the lastChange point. Like log-based debugging, the program state and the call stack are recorded during program execution. We can't halt the program at \textit{lastChange} because we don't know which is the last one until we return to the original breakpoint.

@@

The \textit{lastChange} algorithm is based on program re-execution. \textit{Reproduction point} is the execution point in which first \textit{lastChange} is calld. Like the introductory example, it can be specified by a [conditional] breakpoint. \textit{lastChange} can be called on two different types of values: An object property or a variable value. In JavaScript, every available variable in a frame comes from one of the scopes in the current scope chain. There are five different scope types: global, local, with, catch, closure. We explain each case separately.

@@

Once developer asks for the last change of \texttt{bar.foo} at the reproduction point, debugger calls \texttt{setPropertyChangeHook()} with \texttt{foo} as the property name and re-executes the program. Whenever \texttt{foo} changes and the hook function is called, debugger first calls \texttt{objectID()} on the \texttt{foo} owner object. Then it stores owner id and stack frame locations. The reason we keep object id instead of a reference to the object itself is to let garbage collector destroys dead objects. Whenever the execution reaches the reproduction point it looks at the history of \texttt{foo} changes and finds the last \texttt{foo} change with the same object id as \texttt{bar} id at the reproduction point(Figure ~\ref{fig:foo-changes1}).

\begin{figure}[htp]
\includegraphics[width=.48\textwidth]{6-foo-changes1.jpg}
\caption{The list of \texttt{foo} changes and \texttt{bar} id which identifies the last change of \texttt{bar.foo} in the history of events.}
\label{fig:foo-changes1}
\end{figure}

\subsection{\textit{lastChange} on variable} 
Once developer asks for the last change of variable \texttt{foo} at a reproduction point, debugger first determines the variable's scope as follows: it iterates over the scopes in the scope chain and the first scope which has a variable with the same name is the variable's scope. Consider that in JavaScript, available variables are not resricted to local varialbes, but all variables in outer scopes are also accessible. Among five scope types, global scopes and with scopes-scopes which are created by calling \texttt{with()} function-are regular JavaScript objects. Debugger treats the case where variable's scope is global or with, like \textit{lastChange} on an object property.

A closure scope is created once a function is called. In this case, debugger sets breakpoints on the line the variable is defined and all lines the variable is assigned a new variable (including inner-functions). For example in Figure ~\ref{fig:js-closure}, if the execution is paused at line 24 and the last change of \texttt{var2} is requested by the developer, two breakpoints on lines 22 and 23 will be set, but if the execution is paused at line 19, four breakpoints on lines 11, 12, 16, and 18.

%%% we need a short explanation about scopes
%%% important: think about it: we also need to set a special variable (id) to let us to identify the related functional scope
\begin{figure}[htp]
\begin{verbatim}
10  function parent(var1){
11    var var2;
12    var2 = 1;
13    function myfun(var3){
14      var var4;
15      var4 = {...}
16      var2 = var4;
17      function firstChild(var5){
18        var2 = 5;
19        var4 = 5;
20      }  
21      function secondChild(var6){
22       	var var2;
23        var2 = 5			      
24        var4 = 5;	
25      }
26    }  
23  }    
\end{verbatim}
\caption{Sample JavaScript code demonstrating closure variables.}
\label{fig:js-closure}
\end{figure}

Everytime one of the breakpoints hit a new event is inserted to the list of events. To separate irrelvant event from each other, we use a special variable \texttt{\_scopeId}. If the scope has this variable, the value of this variable is used for the scope id otherwise a new id will be generated and \texttt{\_scopeId} variable is created in the scope. This is specially needed in case of recursive calls. Figure Figure ~\ref{fig:recursive} shows changes to a the same variable \texttt{x} but in different scopes. alling \textit{lastChange(x)} at point B should return point A.

@@

\subsection{\textit{lastChange} on \textit{lastChange}}
Once the developer asks for the first \textit{lastChange} all program state is available to the debugger. But the next \textit{lastChange} calls are asked over collected data. For example, debugger needs object id for the second \textit{lastChange} and this data should be collected. To do so, debugger does a data dependency analysis between points which gives a list of items should be collected at every point.

>>>>

As demonstrated in Figure ~\ref{fig:example-points}, the developer examinated three points of execution. Excepting the first point which debugger pause execution at that point, the second and third points are in the past and are shown by dotted line. These points-the history of the search for the defect-are available through the debugger's interface. On the top of the left panel in Figure ~\ref{fig:example3} there is an opened list which shows all three examinated points. The first one is the breakpoint on line 22, the second one is the point which is the last change of \texttt{myObject.myProperty} before the reaching the breakpoint and finally the last one is the point of execution in which \texttt{myCondition.value} gets the \texttt{undefined} value. Moreover, the source lines related to these points are tagged with red-cycle icons.

Notice that in our example, \textit{lastChange} combines some aspects of breakpoint and of log-based debugging. Like breakpoint debugging, the developer re-executes a live runtime without changing the source and without a special execution environment beyond the debugger. The state of the program memory and the call stack are available at the lastChange point. Like log-based debugging, the program state and the call stack are recorded during program execution. We can't halt the program at \textit{lastChange} because we don't know which is the last one until we return to the original breakpoint. In section 5 we discuss about the possibility of pausing at \textit{lastChange} results.

@@

The \textit{lastChange} algorithm is based on program re-execution. Once developer asks for the \textit{lastChange} of a value, debugger re-executes the program, collects limited data and shows the result. \textit{Reproduction point} is the execution point in which first \textit{lastChange} is calld. Like the introductory example, it can be defined by a [conditional] breakpoint. \textit{lastChange} can be called on two different types of values: An object property or a variable value. In JavaScript, every available variable in a frame comes from one of the scopes in the current scope chain. There are five different scope types: global, local, with, catch, closure. We explain each case separately.

@@

Once developer asks for the last change of \texttt{bar.foo} at the reproduction point, debugger calls \texttt{setPropertyChangeHook()} with \texttt{foo} as the property name and re-executes the program. Whenever \texttt{foo} changes and the hook function is called, debugger first calls \texttt{objectID()} on the \texttt{foo} owner object. Then it stores owner id and stack frame locations. The reason we keep object id instead of a reference to the object itself is to let garbage collector to destroy dead objects. Whenever the execution reaches the reproduction point it looks at the history of \texttt{foo} changes and finds the last \texttt{foo} change with the same object id as \texttt{bar} id at the reproduction point(Figure ~\ref{fig:foo-changes1}).

\begin{figure}[htp]
\includegraphics[width=.48\textwidth]{6-foo-changes1.jpg}
\caption{The list of property \texttt{foo} changes and \texttt{bar} id which identifies the last change of \texttt{bar.foo} in the history of events.}
\label{fig:foo-changes1}
\end{figure}

\subsection{\textit{lastChange} on variable} 
In JavaScript, every frame has a scope chain and every available variable in the frame comes from one of the scopes in the frame's scope chain. Once developer asks for the last change of variable \texttt{foo} at the reproduction point, debugger first determines the variable's scope as follows: it iterates over the scopes in the scope chain and the first scope which has a variable with the same name is the variable's scope. There are five different scope types: local, global, \texttt{with}, closure, \texttt{catch}. We explain each case separately.

\subsubsection{global and \texttt{with} scopes}
Global scope is the the most outer scope in the scope chain and it also refers as global object. This scope is a regular JavaScript object and therefore every global variable is a property of global object. Similar to global scope, \texttt{with} scopes are also regular objects. A \texttt{with} scope is created by a \texttt{with()} block with an object as the parameter. Every property of passed object is available inside the block as a variable. Debugger treats the case where variable's scope is global or \texttt{with}, like \textit{lastChange} on an object property.

\subsubsection{local, closure and \texttt{catch} scopes}
local scope refers to the top scope in the scope chain. Clousure scope refers to the scope which is created once a function is defined. Catch scope is the scope created by try-catch blocks and contains the exception variable. These objects are not necessarily regular JavaScript objects. Therefore to track changes to a variable in these scopes we need a new strategy.

%%% we need a short explanation about scopes
%%% important: think about it: we also need to set a special variable (id) to let us to identify the related functional scope
\begin{figure}[htp]
\begin{verbatim}
10  function parent(arg1){
11    var var1;
12    var1 = 0;
13    function myfun(arg2){
14      var var2;
15      var2 = {...}
16      var1 = var2;
17      function firstChild(arg3){
18        var1 = 1;
19        var2 = 1;
20      }  
21      function secondChild(arg4){
22       	var var1;
23        var1 = 2			      
24        var2 = 2;	
25      }
26    }  
27  }    
\end{verbatim}
\caption{Sample JavaScript code demonstrating closure variables.}
\label{fig:js-closure}
\end{figure}

A variables in these scopes can only be changed by the code which is included in the it's definition block. Therefore it is only enough to take care only about the lines that a variable is assinged a new value. The same fact is also true about \texttt{catch} and local scopes. Debugger puts hooks in on the line the variable is defined and all lines the variable is assigned a new variable (including inner-functions). For example in Figure ~\ref{fig:js-closure}, if the execution is paused at line 24 and the last change of \texttt{var1} is requested by the developer, two hooks on lines 22 and 23 will be set, but if the execution is paused at line 19, four hooks on lines 11, 12, 16, and 18. The reason is that \texttt{var1} in \texttt{firstChild} is a closure variable but in \texttt{secondChild} is a local variable.

@@

\begin{figure}[htp]
\includegraphics[width=.48\textwidth]{7-foo-changes2.jpg}
\caption{The list of variable \texttt{foo} changes and the \texttt{foo}'s scope id at the reproduction point which identifies the answer.}
\label{fig:foo-changes2}
\end{figure}

Everytime one of the hook hit a new event is inserted to the list of events. To separate irrelvant event from each other, we use a special variable \texttt{\_scopeId}. If the scope has this variable, the value of this variable is used for the scope id otherwise a new id will be generated and \texttt{\_scopeId} variable is created by \texttt{eval()} function in the scope. For example evaluating \texttt{var \_scopeId = 50} will create a variable with name \texttt{\_scopeId} and value \texttt{50} in the local scope. Consider that even in the case of closure scopes, \texttt{\_scopeId} is set when the variable first defined and the variable's scope is still local. Figure ~\ref{fig:recursive} demonstrates one case that scope id is necessary for identifying the right answer. In thiss case the same variable \texttt{x} is changed several times but in different scopes due to recursive calls. Without the scope id calling \textit{lastChange(x)} at point C will return point B while the right answer is point A.

\subsection{\textit{lastChange} on \textit{lastChange}}
The algorithm we explained so far needs an enhancement to support longer sequence of \textit{lastChange}s. We explain first the issue and then the solution. Consider the following points:

\begin{center}
\textit{
 point A : the reproduction point \\
 point B : lastChange(A, bar.x) \\
 point C : lastChange(B, baz.y) 
 }
 \end{center}
According the algorithm explained in the previous section none of the points B and C can be identified before reaching point A. Therefore, all \texttt{x} changes for piont B and all \texttt{y} changes for point C are stored. Once the execution reaches point A, and the exact \texttt{bar} id reveals, point B can be identified among the items in the history of \texttt{x} changes. The problem arises in recognizing point C. To recognize point C, we need \texttt{baz} id at point B, but B is a past program state, and \texttt{baz} id has not been collected at this point.

\begin{figure}[htp]
\includegraphics[width=.48\textwidth]{8-lastchange-lastchange.jpg}
\caption{The list of events stored for locating \textit{lastChange} of \texttt{bar.x} (point B) at the reproduction point and \texttt{baz.y} at point B.}
\label{fig:lastchange-lastchange}
\end{figure}

To address this issue, debugger does a dependency analysis between all \textit{lastChange} queries and computes the additional data needs to be collected at each point. For example in the mentioned case, \texttt{baz} id- if it is available- should be collected at all points \texttt{x} changes. Having this additional data, locating point C, using the \texttt{baz} id at point B is straightforward (Figure ~\ref{fig:lastchange-lastchange}).
